Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-27 ms.tex
cutting fat - and lots of it9
cutting fat - and lots of it10

@@

Cooling instability can, however, be prevented if destabilizing
radiative cooling ($\dot{q}_{\rm rad} \propto T^{-2.7}$) is
overwhelmed by other sources of cooling, namely the {\it stabilizing}
term $\propto -q c_{s}^{2} \propto -T$ in the entropy equation
(eq.~[\ref{eq:dsdt}]).  Neglecting radiative cooling to first order,
this term is balanced at the stagnation radius by the external heating
term, $\dot{q}_{\rm heat} = q \tilde{v}_{w}^{2}/2$.  We assess thermal
stability near the stagnation radius by comparing the ratio of
external heating to radiative cooling $|\dot{q}_{\rm rad}|$
(eq.~[\ref{eq:qdot_rad}]),
\begin{align}
\left.\frac{\dot{q}_{\rm heat}}{|\dot{q}_{\rm rad}|}\right|_{\rm r_{\rm s}} \simeq
  \begin{cases}
   630 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.76}v_{500}^{7.2}  &, \text{core}\\
   660 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.48}v_{500}^{5.8}  &, \text{cusp},     
  \end{cases}
  \label{eq:cooling2}
\end{align}
where we have used equations (\ref{eq:rhors}) and (\ref{eq:rhostarrs})
for the gas and stellar densities at the stagnation radius,
respectively.  We have approximated the cooling function for $T <
2\times 10^{7}$ K as $\Lambda(T) = 1.1 \times 10^{-22} \left(T/10^6
  \text{K}\right)^{-0.7} $erg cm$^3 $s$^{-1}$, which assumes solar
metallicity gas (\citealt{Draine:2011a}; his Fig. 34.1).

@@

Based on our numerical results (\S \ref{sec:numerical}) and the work of \citet{McCourt+12} we define thermally stable flows according to the criterion $\dot{q}_{\rm heat} > 10|\dot{q}_{\rm rad}|$ ($t_{\rm
  cool} > 10t_{\rm ff}$) being satisfied near the stagnation radius.  This condition translates into a
critical minimum heating rate
\begin{align}
v_{w} > v_{\rm TI} \simeq
  {\rm max}\begin{cases}
   280 \eta_{0.02}^{0.14} M_{\bullet,8}^{0.11}\,{\rm km\,s^{-1}}  &, \text{core}\\
   240 \eta_{0.02}^{0.17} M_{\bullet,8}^{0.08}\,{\rm km\,s^{-1}}   &, \text{cusp}  \\
\approx \sigma_0(r_{b}/r_{\rm inf})^{0.5(1-\Gamma)} &, \zeta \ll \zeta_c, 
  \end{cases}
  \label{eq:cooling3}
\end{align}
where the final condition corresponds to the low wind heating limit
($\zeta \ll \zeta_c$; eq.~[\ref{eq:zetacrit}]), for which our
numerical and analytic results show that the stagnation radius moves
out to the break radius $r_b$; the quasi-hydrostatic structure that
develops in this case greatly increases the gas density, resulting in local
thermal instability.

Note that in practice when $v_{w}<v_{\rm TI}$, a cooling flow may
develop rather than a true local thermal instability: in the former
case the stagnation radius $\rs$ may move to large radii, greatly
increasing the density in the inner part of the flow, which increases
cooling and creates a large inflow of cold gas towards the center of
the galaxy.  In this paper we do not attempt to distinguish under what
conditions a global cooling flow would develop vs. a local thermal
instability.

@@

@@

Our analytic estimates accurately reproduce the numerical results in
the high heating limit $\zeta \gg 1$ ($v_w \gg \sigma_0$; $r_{\rm s}
\lesssim r_{\rm inf}$).  However, for low heating the stagnation
radius diverges above the analytic estimate, approaching the stellar
break radius $r_b \gg r_{\rm inf}$.  This divergence occurs
approximately when $\zeta < \zeta_c \propto (r_b/r_{\rm
  soi})^{0.5(\Gamma-1)}$ (eq.~[\ref{eq:zetacrit}]), which coincides
with the Bernoulli parameter of the gas decreasing below unity near
the stagnation radius (Appendix \ref{app:be}).  Thus, for small values
of the heating rate (small $\zeta$) the location of the stagnation
radius will vary strongly with the break radius. This explains the
behavior of the three vertically aligned green squares. These are three
cusp ($\Gamma=0.8$) galaxies with t$\Mbh=10^8 \Msun$ and $v_w=300$ km s$^{-1}$
but with different break radii ($\rb=$ 25, 50, and 100 pc from top to
bottom). This divergence of the stagnation radius to large radii
occurs at a higher value of $\zeta$ in core galaxies ($\Gamma = 0.1$),
explaining the behavior of the two core galaxies shown in
Fig.~\ref{fig:stag} as vertically aligned orange triangles.  
%These
%solutions correspond to the same black hole mass ($\Mbh = 10^8 \Msun$)
%and heating rate ($v_w = 600$ km s$^{-1}$) but assume different break
%radii of $r_{b} =$ 25, 50 pc (from top to bottom).

>>>>

\begin{figure}
  \includegraphics[width=\columnwidth]{TI.pdf}
  \caption{
 }
\label{fig:TI}
\end{figure}

@@

Cooling instability can, however, be prevented if destabilizing
radiative cooling ($\dot{q}_{\rm rad} \propto T^{-2.7}$) is
overwhelmed by other sources of cooling, namely the {\it stabilizing}
term $\propto -q c_{s}^{2} \propto -T$ in the entropy equation
(eq.~[\ref{eq:dsdt}]).  Neglecting radiative cooling to first order,
this term is balanced at the stagnation radius by the external heating
term, $\dot{q}_{\rm heat} = q \tilde{v}_{w}^{2}/2$.  One can therefore assess thermal
stability by comparing the ratio of
external heating to radiative cooling $|\dot{q}_{\rm rad}|$
(eq.~[\ref{eq:qdot_rad}]),
\begin{align}
\left.\frac{\dot{q}_{\rm heat}}{|\dot{q}_{\rm rad}|}\right|_{\rm r_{\rm s}} \simeq
  \begin{cases}
   630 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.76}v_{500}^{7.2}  &, \text{core}\\
   660 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.48}v_{500}^{5.8}  &, \text{cusp},     
  \end{cases}
  \label{eq:cooling2}
\end{align}
where we have used equations (\ref{eq:rhors}) and (\ref{eq:rhostarrs})
for the gas and stellar densities at the stagnation radius,
respectively.  We have approximated the cooling function for $T <
2\times 10^{7}$ K as $\Lambda(T) = 1.1 \times 10^{-22} \left(T/10^6
  \text{K}\right)^{-0.7} $erg cm$^3 $s$^{-1}$, which assumes solar
metallicity gas (\citealt{Draine:2011a}; his Fig. 34.1).

@@

Based on our numerical results (\S \ref{sec:numerical}) and the work of \citet{McCourt+12} we define thermally stable flows according to the criterion $\dot{q}_{\rm heat} > 10|\dot{q}_{\rm rad}|$ ($t_{\rm
  cool} > 10t_{\rm ff}$) being satisfied near the stagnation radius.  This condition translates into a
critical minimum heating rate
\begin{align}
v_{w} > v_{\rm TI} \simeq
  {\rm max}\begin{cases}
   280 \eta_{0.02}^{0.14} M_{\bullet,8}^{0.11}\,{\rm km\,s^{-1}}  &, \text{core}\\
   240 \eta_{0.02}^{0.17} M_{\bullet,8}^{0.08}\,{\rm km\,s^{-1}}   &, \text{cusp}  \\
%\approx \sigma_0(r_{b}/r_{\rm inf})^{0.5(1-\Gamma)} &, \zeta \ll \zeta_c. 
  \end{cases}
  \label{eq:cooling3}
\end{align}
Equations (\ref{eq:cooling2}) and (\ref{eq:cooling3}) are derived using expressions for the stagnation radius and gas density in the high heating limit of $\zeta > \zeta_c = (r_{b}/r_{\rm inf})^{0.5(1-\Gamma)}$ (eq.~[\ref{eq:zetacrit}]).  However, for $\zeta < \zeta_c$, the stagnation radius diverges to the break radius $r_b$ (Appendix \ref{app:be}).  The quasi-hydrostatic structure that results in this case greatly increases the gas density, which in practice we find renders the flow susceptible to thermal runaway, even if $v_{w} > v_{\rm TI}$ according to equation (\ref{eq:cooling3}).  In other words, the true condition for thermal instability  can be written
\be
v_w > {\rm max}[v_{\rm TI},\sigma_0\{1-(r_{b}/r_{\rm inf})^{0.5(1-\Gamma)}\}]
\ee

Figure \ref{fig:TI} shows $v_{\rm TI}(M_{\bullet})$ for different mass input parameters $\eta$, as well as the $\zeta > \zeta_c$ criterion, shown separately for cusp and core galaxies.  We calculate $\zeta_c$ using the empirical correlation for
core galaxies $r_{b}/r_{\rm inf} \approx 7.6M_{\bullet,8}^{-0.2}$ which
follows from equation (\ref{eq:rsoi}) and the $r_b-M_{\bullet}$
correlation derived from the results of \citet{LauerFaber+:2007a}.
For cusp galaxies we see that the $v_w \gtrsim v_{\rm TI}$ criterion is generally more strigent, while for core galaxies the $\zeta > \zeta_c$ criterion is more strigent for core galaxies, except for very high mass loss rates $\eta > 10$.

@@

Note that in practice  what we describe as "thermal instability" may simply indicate the transition of an inflow-outflow solution to a global cooling flow, instead of a true local thermal instability.  In the former case the stagnation radius diverges to large radii, increasing the density in the inner parts of the flow, which increases cooling and creates a large inflow of cold gas towards the nucleus.  A true thermal instability would cause parts of the hot ISM to condense into cold clouds.  In this paper we do not distinguish between these possibilities, although both are likely present at some level.  Note that if the CNM were to somehow ``regulate" itself to a state of local marginal thermal instability (as has previously been invoked on cluster scales; e.g., \citealt{Voit+15}), then equation (\ref{eq:Mdotmax}) could reflect the assocated mass ``fall-out" rate and concomitant star formation rate.

@@

Our analytic estimates accurately reproduce the numerical results in
the high heating limit $\zeta \gg 1$ ($v_w \gg \sigma_0$; $r_{\rm s}
\lesssim r_{\rm inf}$).  However, for low heating the stagnation
radius diverges above the analytic estimate, approaching the stellar
break radius $r_b \gg r_{\rm inf}$.  This divergence occurs
approximately when $\zeta < \zeta_c \propto (r_b/r_{\rm
  soi})^{0.5(\Gamma-1)}$ (eq.~[\ref{eq:zetacrit}]), which coincides
with the Bernoulli parameter of the gas decreasing below unity near
the stagnation radius (Appendix \ref{app:be}).  Thus, for small values
of the heating rate (small $\zeta$) the location of the stagnation
radius will vary strongly with the break radius. This explains the
behavior of the three vertically aligned green squares. These are three
cusp ($\Gamma=0.8$) galaxies with $\Mbh=10^8 \Msun$ and $v_w=300$ km s$^{-1}$
but with different break radii ($\rb=$ 25, 50, and 100 pc from top to
bottom). This divergence of the stagnation radius to large radii
occurs at a higher value of $\zeta$ in core galaxies ($\Gamma = 0.1$),
explaining the behavior of the two core galaxies shown in
Fig.~\ref{fig:stag} as vertically aligned orange triangles.  
%These
%solutions correspond to the same black hole mass ($\Mbh = 10^8 \Msun$)
%and heating rate ($v_w = 600$ km s$^{-1}$) but assume different break
%radii of $r_{b} =$ 25, 50 pc (from top to bottom).
